Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-20 mr-review.Rmd
added notes for conclusion section
added text about zeng 2019 birth weight

In the short time since publication of @verbanck_detection_2018, researchers have recognized two 
types of horizontal pleiotropy [@yuan_likelihood_2021]. The first occurs via 
exposure-independent paths. The resulting horizontal pleiotropic effects are independent of the
SNP-exposure relationships. The second type of horizontal pleiotropy manifests when 
in the presence of unobserved exposure-outcome confounding. It induces correlation between 
horizontal pleiotropic effects and SNP-exposure effects. Both types of horizontal pleiotropy 
violate standard MR modeling assumptions and can bias causal effect estimates and can increase false 
discoveries [@yuan_likelihood_2021].
Early MR analyses avoided confounding from horizontal pleiotropy by discarding 
instrumental SNPs that might be associated with the outcome 
[@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-1;@zeng_causal_2019-2].
More recent methods have attempted to model horizontal pleiotropy 
[@morrison_mendelian_2020;@yuan_testing_2020;@yuan_likelihood_2021]. CAUSE [@morrison_mendelian_2020] and 
MRMix [@qi_mendelian_2019] both use a mixture of normal distributions to control for both 
types of horizontal pleiotropy. Modeling both types of horizontal pleiotropy is particularly 
challenging because the MR model likelihood often involves an integral that can't be solved analytically. 
Because of this issue with the model likelihood, both MRMix and CAUSE use other, non-likelihood-based methods for inference.
CAUSE compares the out-of-sample prediction accuracies for two competing models that differ in inclusion of 
the causal effect parameter [@morrison_mendelian_2020]. Specifically, CAUSE calculates the expected pointwise log posterior density 
between the two models. 
@yuan_likelihood_2021 found that MRMix is not robust to misspecification of SNP effect sizes and often is 
biased. CAUSE, they found, yields overly conservative p-values [@yuan_likelihood_2021].

@@

@@

1. Relatively small proportion of SNPs have nonzero effects on the exposure  
1. A relatively small proportion of SNPs demonstrate horizontal pleiotropy  
1. The chosen instrumental SNPs are more likely to display horizontal pleiotropy than are non-instrumental SNPs  
1. Those SNPs that display horizontal pleiotropy are more likely to demonstrate uncorrelated horizontal pleiotropy than correlated horizontal pleiotropy.

@@

@zeng_birth_2019

@@

>>>>

In the short time since publication of @verbanck_detection_2018, researchers have recognized two 
types of horizontal pleiotropy [@yuan_likelihood_2021]. The first occurs via 
exposure-independent paths. The resulting horizontal pleiotropic effects are independent of the
SNP-exposure relationships. The second type of horizontal pleiotropy manifests when 
in the presence of unobserved exposure-outcome confounding. It induces correlation between 
horizontal pleiotropic effects and SNP-exposure effects. Both types of horizontal pleiotropy 
violate standard MR modeling assumptions and can bias causal effect estimates and can increase false 
discoveries [@yuan_likelihood_2021].
Early MR analyses avoided confounding from horizontal pleiotropy by discarding 
instrumental SNPs that might be associated with the outcome 
[@verbanck_detection_2018;@zhu_causal_2018;@zeng_causal_2019-1;@zeng_causal_2019-2].
More recent methods have attempted to model horizontal pleiotropy 
[@morrison_mendelian_2020;@yuan_testing_2020;@yuan_likelihood_2021]. CAUSE [@morrison_mendelian_2020] and 
MRMix [@qi_mendelian_2019] both use a mixture of normal distributions to control for both 
types of horizontal pleiotropy. Modeling both types of horizontal pleiotropy is particularly 
challenging because the MR model likelihood often involves an integral that can't be solved analytically. 
Because of this issue with the model likelihood, both MRMix and CAUSE use other, non-likelihood-based methods for inference.

@@

@yuan_likelihood_2021 found that MRMix is not robust to misspecification of SNP effect sizes and often is 
biased. CAUSE, they found, yields overly conservative p-values [@yuan_likelihood_2021].

@@

1. Relatively small proportion of SNPs have nonz
ero effects on the exposure  
1. A relatively small proportion of SNPs demonstrate horizontal pleiotropy  
1. The chosen instrumental SNPs are more likely to display horizontal pleiotropy than are non-instrumental SNPs  
1. Those SNPs that display horizontal pleiotropy are more likely to demonstrate uncorrelated horizontal pleiotropy than correlated horizontal pleiotropy.

@@

Birth weight is associated with childhood asthma, but whether there is a causal relationship 
between the two remains unclear [@brooks_impact_2001]. To investigate this possibility, 
@zeng_birth_2019 performed a MR analysis to assess the causal effect of birth weight on adult 
asthma. They used data from two GWAS, with 140,000 subjects in the birth weight study and 62,000 
subjects in the adult asthma study. With both the genetic risk score method and the 
inverse variance-weighted method, @zeng_birth_2019 found no evidence of a causal effect of birth
weight on adult asthma. Sensitivity analysis results offered reassurance that the 
finding of no causal effect is robust. MR-PRESSO and MR-Egger found no evidence for bias from 
SNP instrument outliers or horizontal pleiotropy. @zeng_birth_2019 concludes that confounding
in other studies may explain the discrepancy between their findings and others' conclusions.

@@

Other biobanks?

EMR & GWAS?
